---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE, show_col_types = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=4)
```
# Data preprocessing
```{r}
library(dplyr)
library(readr)

# Read the data
nba_data <- read_csv("nba_final.csv")

# Clean the data by removing observations with NA entries
nba_data_final <- nba_data %>%
  filter(!is.na(Age), 
         !is.na(PTS),
         !is.na(AST),
         !is.na(TRB),
         !is.na(Pos1),
         !is.na(mean_views),
         Salary >0)

nba_data_final <- nba_data_final %>%
  mutate(Pos1 = factor(Pos1))
```
# EDA
```{r}
# histogram of Salary
hist(nba_data_final$Salary, breaks = 30)
# boxplot of Salary
boxplot(nba_data_final$Salary, xlab = "Salary (USD)", horizontal = TRUE)

# Salary vs each predictor
par(mfrow = c(2, 3),
mar = c(4, 4, 2, 1))

plot(nba_data_final$Age, nba_data_final$Salary,
     main = "Salary vs Age",
     xlab = "Age",
     ylab = "Salary (USD)",
     pch = 16,           # Solid circle
     col = "steelblue")

# Add linear regression line
abline(lm(Salary ~ Age, data = nba_data_final), col = "red", lwd = 2)

# Scatter plot of Salary vs PTS
plot(nba_data_final$PTS, nba_data_final$Salary,
     main = "Salary vs Points (PTS)",
     xlab = "Points (PTS)",
     ylab = "Salary (USD)",
     pch = 16,
     col = "darkgreen")

# Add linear regression line
abline(lm(Salary ~ PTS, data = nba_data_final), col = "red", lwd = 2)

# Scatter plot of Salary vs AST
plot(nba_data_final$AST, nba_data_final$Salary,
     main = "Salary vs Assists (AST)",
     xlab = "Assists (AST)",
     ylab = "Salary (USD)",
     pch = 16,
     col = "purple")

# Add linear regression line
abline(lm(Salary ~ AST, data = nba_data_final), col = "red", lwd = 2)

# Scatter plot of Salary vs TRB
plot(nba_data_final$TRB, nba_data_final$Salary,
     main = "Salary vs Rebounds (TRB)",
     xlab = "Rebounds (TRB)",
     ylab = "Salary (USD)",
     pch = 16,
     col = "orange")

# Add linear regression line
abline(lm(Salary ~ TRB, data = nba_data_final), col = "red", lwd = 2)

# Scatter plot of Salary vs Mean Views
plot(nba_data_final$mean_views, nba_data_final$Salary,
     main = "Salary vs Mean Views",
     xlab = "Mean Views",
     ylab = "Salary (USD)",
     pch = 16,
     col = "brown")

# Add linear regression line
abline(lm(Salary ~ mean_views, data = nba_data_final), col = "red", lwd = 2)

# Boxplot of Salary by Position
boxplot(Salary ~ Pos1, data = nba_data_final,
        main = "Salary by Position",
        xlab = "Position",
        ylab = "Salary (USD)",
        col = "lightcoral",
        border = "darkred")
```
# Checking multicolinearity
```{r}
predictor_fit <- lm(PTS ~ Age + AST + TRB + Pos1 + mean_views, data = nba_data_final)
summary(predictor_fit)
```
# Fit the multiple regression model
```{r}
nba_model <- lm(Salary ~ Age * PTS + AST + TRB + Pos1 * mean_views, data = nba_data_final)

# View the summary of the model
summary(nba_model)
```
# Diagnostic plots
```{r}
par(mfrow = c(2, 2))
plot(nba_model, which = c(1, 2, 3, 4))

# Closer look at diagnostic plots
par(mfrow = c(1,1))
residuals <- resid(nba_model)
fitted_values <- fitted(nba_model)

# Residuals vs Fitted
plot(fitted_values, residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")  # Horizontal line at zero

# Normal Q-Q Plot
qqnorm(residuals, main = "Normal Q-Q Plot")
qqline(residuals, col = "red")

# Scale-Location Plot
sqrt_std_resid <- sqrt(abs(residuals))  # Square root of standardized residuals
plot(fitted_values, sqrt_std_resid,
     xlab = "Fitted Values",
     ylab = "Square Root of |Residuals|",
     main = "Scale-Location Plot")
lines(loess.smooth(fitted_values, sqrt_std_resid), col = "red", lwd = 2)
```
# residuals vs each predictor
```{r}
predictors <- nba_data_final[, c("Age", "PTS", "AST", "TRB", "mean_views", "Pos1")]  # Select predictors
par(mfrow = c(2, 3))
residuals <- resid(nba_model)
for (predictor in names(predictors)) {
  plot(predictors[[predictor]], residuals, 
       xlab = predictor, 
       ylab = "Residuals", 
       main = paste("Residuals vs", predictor),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
}
```
# Transformations
```{r}
# Perform Box-Cox transformation
library(MASS)

# Apply Box-Cox transformation to find optimal lambda
boxcox_fit <- boxcox(nba_model, lambda = seq(-2, 2, 0.1))
lambda_optimal <- boxcox_fit$x[which.max(boxcox_fit$y)]  # Extract the optimal lambda
lambda_optimal

# Apply the transformation with the optimal lambda
lambda <- 0.3
if (lambda == 0) {
  nba_data_final$Salary_transformed <- log(nba_data_final$Salary)
} else {
  nba_data_final$Salary_transformed <- (nba_data_final$Salary^lambda - 1) / lambda
}

# transform AST
nba_data_final$Age_sq <- nba_data_final$Age^2

# transform mean_views
nba_data_final$log_mean_views <- log(nba_data_final$mean_views)

# Fit the model with the transformed variables
transformed_fit <- lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 * log_mean_views, 
                      data = nba_data_final)

# Summary of the transformed model
summary(transformed_fit)
```
# Verify assumptions for transformed fit
```{r}
# Diagnostic plots
par(mfrow = c(2, 2))
plot(transformed_fit, which = c(1, 2, 3, 4))

# Closer look at diagnostic plots
par(mfrow = c(1,1))
residuals <- resid(transformed_fit)
fitted_values <- fitted(transformed_fit)
sresidual_values <- rstandard(transformed_fit)

# Residuals vs Fitted
plot(fitted_values, residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")  # Horizontal line at zero

# Normal Q-Q Plot
qqnorm(residuals, main = "Normal Q-Q Plot")
qqline(residuals, col = "red")

# Scale-Location Plot
sqrt_std_resid <- sqrt(abs(residuals))  # Square root of standardized residuals
plot(fitted_values, sqrt_std_resid,
     xlab = "Fitted Values",
     ylab = "Square Root of |Residuals|",
     main = "Scale-Location Plot")
lines(loess.smooth(fitted_values, sqrt_std_resid), col = "red", lwd = 2)

hist(sresidual_values,
     main = "Standardized residuals histogram",
     xlab = "Standardized residuals")

plot(nba_data_final[, c(46, 4, 47, 26, 27, 32, 48)], col="cadetblue")
```
# Comparison between preliminary and transformed model
```{r}
par(mfrow = c(1,2))

# Closer look at diagnostic plots

residuals <- resid(nba_model)
fitted_values <- fitted(nba_model)

Tresiduals <- resid(transformed_fit)
Tfitted_values <- fitted(transformed_fit)
Tsresidual_values <- rstandard(transformed_fit)


# Residuals vs Fitted
plot(fitted_values, residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Preliminary Residuals vs Fitted")
abline(h = 0, col = "red")  # Horizontal line at zero
# Residuals vs Fitted
plot(Tfitted_values, Tresiduals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Transformed Residuals vs Fitted")
abline(h = 0, col = "red")  # Horizontal line at zero
# Normal Q-Q Plot

qqnorm(residuals, main = "Preliminary Normal Q-Q Plot")
qqline(residuals, col = "red")



# Normal Q-Q Plot
qqnorm(Tresiduals, main = "Transformed Normal Q-Q Plot")
qqline(Tresiduals, col = "red")

# residuals vs Age
plot(nba_data_final$Age, residuals, 
       xlab = "Age", 
       ylab = "Residuals", 
       main = paste("Residuals vs", "Age"),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
# residuals vs Age_sq
plot(nba_data_final$Age_sq, Tresiduals, 
       xlab = "Age_sq", 
       ylab = "Residuals", 
       main = paste("Transformed Residuals vs", "Age_sq"),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
  
  # residuals vs mean_views
plot(nba_data_final$mean_views, residuals, 
       xlab = "mean_views", 
       ylab = "Residuals", 
       main = paste("Residuals vs", "mean_views"),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
# residuals vs log_mean_views
plot(nba_data_final$log_mean_views, Tresiduals, 
       xlab = "log_mean_views", 
       ylab = "Residuals", 
       main = paste("Transformed Residuals vs log_mean_views"),
       col = "cadetblue", pch = 16, cex.main = 0.9)
  abline(h = 0, col = "red", lwd = 2)
```
# residuals vs each predictor
```{r}
predictors <- nba_data_final[, c("Age_sq", "PTS", "AST", "TRB", "log_mean_views", "Pos1")]  # Select predictors
par(mfrow = c(2, 3))

residuals <- resid(transformed_fit)

for (predictor in names(predictors)) {
  plot(predictors[[predictor]], residuals, 
       xlab = predictor, 
       ylab = "Residuals", 
       main = paste("Residuals vs", predictor),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
}
```
# full F-test
```{r}
# Perform ANOVA to compare with the null model
fit_null <- lm(Salary_transformed ~ 1, data = nba_data_final)  # Null model with only intercept
anova(fit_null, transformed_fit)
```
# Partial F-tests
```{r}
# dropping Age_Sq:PTS
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 * log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq + PTS + AST + TRB + Pos1 * log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping Pos1:log_mean_views
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 * log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # not significant
```
```{r}
# dropping Pos1
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping log_mean_views
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping TRB
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + Pos1 + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping AST
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ Age_sq * PTS + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping Age_sq
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
```{r}
# dropping PTS
full_fit = lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
reduced_fit = lm(Salary_transformed ~ PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
anova(reduced_fit, full_fit) # is significant
```
# Final model
```{r}
final_model <- lm(Salary_transformed ~ Age_sq * PTS + AST + TRB + Pos1 + log_mean_views, 
                      data = nba_data_final)
summary(final_model)
```
# Verifying assumptions for final model
```{r}
par(mfrow = c(2, 2))
plot(final_model, which = c(1, 2, 3, 4))
```
# residuals vs each predictor
```{r}
predictors <- nba_data_final[, c("Age_sq", "PTS", "AST", "TRB", "Pos1", "log_mean_views")]  # Select predictors
par(mfrow = c(2, 3))

residuals <- resid(final_model)

for (predictor in names(predictors)) {
  plot(predictors[[predictor]], residuals, 
       xlab = predictor, 
       ylab = "Residuals", 
       main = paste("Residuals vs", predictor),
       col = "cadetblue", pch = 16)
  abline(h = 0, col = "red", lwd = 2)
}
```
# VIF
```{r}
library(car)
vif(final_model)
```
# Outliers, influential points, leverage points
```{r}
# outliers
plot(final_model, which = c(3))
std_residuals <- rstandard(final_model)
outliers <- which(abs(std_residuals) > 4)
outliers
nba_data_final[538,]
# high leverage points
hat <- hatvalues(final_model)
threshold <- 2 * mean(hat)
plot(hat)
abline(h = threshold, col = "red", lty = 2)
# influential points
plot(final_model, which = c(4))
nba_data_final[538,]
nba_data_final[928,]
nba_data_final[1066,]
```
# Inference
```{r}
confidence_intervals <- confint(final_model)
print(confidence_intervals)
```